Efficient ancestry and mutation simulation with msprime 1.0

### 2021-08-19 paper.tex

One of the key reasons for \msprime's substantial performance advantage
over other simulators~\citep{kelleher2016efficient}
is its use of the ``succinct tree sequence''
data structure to represent simulation results.
The succinct tree sequence (usually abbreviated to ``tree sequence'')
was introduced by \cite{kelleher2016efficient}
to concisely encode genetic ancestry and sequence variation
and was originally implemented as part of \msprime.
We subsequently extracted the core tree sequence functionality
from \msprime\ to create the \tskit\ library,
which provides a large suite of tools for processing genetic ancestry
and variation data via APIs
in the Python and C languages~\citep{tskit2021tskit}.
While a full discussion of tree sequences and the capabilities
of \tskit\ is beyond the scope of this article, we summarise
some aspects that are important for simulation.

@@

\subsection*{Integration with forward simulators}
A unique feature of \msprime\ is its ability to simulate
genetic ancestries by extending an existing partial
genetic ancestry. Given a tree sequence that
is complete up until time $t$ ago as input
(where marginal trees may or may not have fully coalesced),
\msprime\ can efficiently obtain the segments of ancestral material
present at this time, and then run the simulation backwards in time from there.
This allows a simulated ancestry to be produced by any
number of different processes across disjoint time slices.
In practice this feature is used to ``complete''
forwards-time ancestry simulations~\citep{kelleher2018efficient}
that may have not fully coalesced. This process
(``recapitation'') can be orders of magnitude faster than
the standard approach of neutral burn-in; see
\cite{haller2018tree} for more details and examples.
This interoperability between simulators, where a partial ancestry
simulation produced by \SLiM~\citep{haller2019slim}
or \fwdpy~\citep{thornton2014cpp} can be picked up and completed
by another simulator, with complete information
retained---at scale---is unprecedented.
This interoperability indicates an opportunity for
other forward genetic simulators ~\citep[e.g.][]{Gaynor2021AlphaSimR}
to leverage the tree sequence data format and associated tools.

>>>>

One of the key reasons for \msprime's substantial performance advantage
over other simulators~\citep{kelleher2016efficient}
is its use of the ``succinct tree sequence''
data structure to represent simulation results.
The succinct tree sequence (usually abbreviated to ``tree sequence'')
was introduced by \cite{kelleher2016efficient}
to concisely encode genetic ancestry and sequence variation
and was originally implemented as part of \msprime.
We subsequently extracted the core tree sequence functionality
from \msprime\ to create the \tskit\ library,
which provides a large suite of tools for processing genetic ancestry
and variation data via APIs
in the Python and C languages~\citep{tskit2021tskit}.
The availability of \tskit\ as a liberally licensed (MIT) open source
toolkit has enabled several other
projects~\citep[e.g.][]{kelleher2019inferring,haller2019slim,wohns2021unified,
de2021geonomics} to take advantage
of the same efficient data structures used in \msprime,
and we hope that many more will follow.
While a full discussion of tree sequences and the capabilities
of \tskit\ is beyond the scope of this article, we summarise
some aspects that are important for simulation.

@@

\subsection*{Integration with forward simulators}
A unique feature of \msprime\ is its ability to simulate
genetic ancestries by extending an existing partial
genetic ancestry. Given a tree sequence that
is complete up until time $t$ ago as input
(where marginal trees may or may not have fully coalesced),
\msprime\ can efficiently obtain the segments of ancestral material
present at this time, and then run the simulation backwards in time from there.
This allows a simulated ancestry to be produced by any
number of different processes across disjoint time slices.
In practice this feature is used to ``complete''
forwards-time ancestry simulations~\citep{kelleher2018efficient}
that may have not fully coalesced. This process
(``recapitation'') can be orders of magnitude faster than
the standard approach of neutral burn-in; see
\cite{haller2018tree} for more details and examples.
This interoperability between simulators, where a partial ancestry
simulation produced by \SLiM~\citep{haller2019slim}
or \fwdpy~\citep{thornton2014cpp} can be picked up and completed
by another simulator, with complete information
retained---at scale---is unprecedented.
There may be an opportunity for
other forward genetic simulators~\citep[e.g.][]{gaynor2021alphasimr}
to leverage the tree sequence data format and associated tools.
